\chapter{Constraint Gen library}

Constraint gen library is designed to generate Points-to constraints from the
LLVM Intermediate Code. In order to generate the constraints, it classifies the
various LLVM IR instructions into four basic pointer assignments as described 
in Section 2.2. This library considers only those LLVM IR statements which are
relevant for \textit{Points-to Analysis} i.e. only those instructions in which 
there is 
read/write to a pointer. This library is completely context and flow 
insensitive. So given any instructions in any order, it will sure generate 
constraint. So, if someone tries to do a forward flow analysis or backward flow 
analysis, or Context sensitive or insensitive, this library will be helpful for 
every case.


\section{Design}
The design of this library is motivated by the fact that most of the points-to 
analysis depends on the pointers used in the source code. So it is designed 
such that it can cover the requirements of the most of the Points-to analyses.

All the information about a variable is stored in a Points-to Node 
(PointsToNode in implementation). Internally this library creates a mapping 
between those PointsToNode and LLVM's Value (every variable or instruction is 
represented by this Value). Each Points-to node is assigned a unique id. This 
id is an unsigned integer. Each constraint represents a single assignment in 
the IR. Section 2.3 discusses more about implementation level details 
of a single constraint.

This library tries to simplify the complex LLVM IR statements into four basic 
pointer assignments. Section 2.2 discusses about the classification of LLVM IR 
statements.

\section{Classification of LLVM Instructions}
Following four basic categories are used to classify the relevant LLVM 
Instructions:

\begin{itemize}
    \item Address Of Instructions, e.g. a = \&b
    \item Load Instructions, e.g.  a = *b
    \item Store Instructions, e.g. *a = b
    \item Copy Instructions, e.g. a = b
\end{itemize}

\subsection{Address Of Instructions}
Address of instructions are those instructions which refers to the assignment 
in which address of some variable is assigned to some other variable/pointer.
Address of instruction corresponds to following standard C statement,
    
\begin{lstlisting}
    a = &b
\end{lstlisting}

In LLVM, this is satisfied by alloca and getelementptr instruction.
Alloca instruction allocates memory in the stack and returns a pointer to it.
Whereas, getelementptr instruction takes a base address and computes the address
of element relative to base address using the provided arguments.
Example,

\begin{lstlisting}
%a = alloca i32, i32 4
%b = getelementptr i32, i32* %a, i32 1   ;select second element                  
                       
\end{lstlisting}

\noindent
First statement is equivalent to, (here b is memory being allocated in stack)

\begin{lstlisting}
    a = &b
\end{lstlisting}

\noindent
And the second statement is equivalent to,
\begin{lstlisting}
    b = &a[1]
\end{lstlisting}

\noindent
Following statements are classified as address of instructions
\begin{itemize}
    \item alloca
    \item getelementptr
\end{itemize}

\subsection{Load Instructions}
Load instructions are those assignments in which the address stored at the 
location pointed by a pointer is assigned to another pointer. So, we are 
basically dereferencing the pointer which points to another pointer and address 
stored in that pointer is assigned to other variable/pointer.
load instruction in LLVM can be used to represent following C statement,
\begin{lstlisting}
    a = *b
\end{lstlisting}

\noindent
Load instruction in LLVM takes an address and reads memory from that address,
e.g.,

\begin{lstlisting}
    %a = load i32*, i32** %b
\end{lstlisting}

\noindent
Here a is a pointer having address of memory from which load will read an i32 
value.
\noindent
Instruction that satisfies this type is,
\begin{itemize}
 \item load
\end{itemize}


\subsection{Store Instructions}
Store instructions are those assignments in which address stored in a pointer 
is assigned to the location pointed by another pointer.
store instruction in LLVM is equivalent to following C statement,
\begin{lstlisting}
    *a = b
\end{lstlisting}

\noindent
Store instruction in LLVM writes to the memory at address provided as second
argument. First argument is the value which has to be written.
Example,
\begin{lstlisting}
    store i32* %b, i32** %a
\end{lstlisting}

\noindent
Constraint-gen library ignores those store statements which do not write address
to memory or which involves writing non pointer values to the memory. So the
following store statements in the IR will be ignored,

\begin{lstlisting}
    store i32 10, i32* %a
\end{lstlisting}

\noindent
Instruction that satisfies this type is,
\begin{itemize}
 \item store
\end{itemize}


\subsection{Copy Instructions}
Copy instructions are those instructions which involve copying of the address 
from one SSA variable to another. For e.g. in phi instruction
\begin{lstlisting}
    %b = phi i32* [ %a, %one ], [ %c, %two ]
\end{lstlisting}

\noindent
It is equivalent to the following C statement,
\begin{lstlisting}
    a = b;
\end{lstlisting}

\noindent
Following instructions satisifies this constraint,
\begin{itemize}
    \item phi
    \item select
    \item extractvalue
    \item insertvalue
\end{itemize}

More details about the constraint being generated for each instruction are
mentioned in next section.

\section{Constraints and their Generation}
Constraints are the information about the pointers assignemnts used in a 
particular IR statement. A single constraint refers to a single possible 
assignment. This assignment can be one of above four type. Each constraint has 
the following properties,

\begin{itemize}
    \item LHS node id
    \item RHS node id
    \item LLVM statement
    \item Type of the constraint, i.e. address of, or load, or store, or copy
\end{itemize}

\noindent
Both LHS and RHS node consists of atleast the following information. 
(Implementation details are discussed in Section 2.4).

\begin{itemize}
    \item the corresponding llvm::Value
    \item what other nodes are being used (See section ??)
    \item Type information, whether it is a struct or an array or a normal 
pointer
    \item A unique identifier
\end{itemize}

\subsection{Generation of constraints}
For each statement type, respective method is used to extract constraint from
that statement type. Constraints are generated for only following instructions.
Other instructions are irrevelant since they do not use pointers or involve
assignments that may affect pointers.

\begin{itemize}
    \item alloca instruction
    \item load instruction
    \item store instruction
    \item getelementptr instruction
    \item phi instruction
    \item bitcast instruction
    \item select instruction
    \item extractvalue instruction
    \item insertvalue instruction
\end{itemize}

\subsubsection{alloca instruction}
Alloca instruction falls under the category of address-of instruction. Alloca 
instruction allocates the memory in stack taking the type and number of 
elements as arguments. It returns the pointer to that memory. 
Example,

\begin{lstlisting}
    %a = alloca i32
\end{lstlisting}
\noindent
Above instruction allocates the memory for i32 value and returns a pointer to 
it. \%a is of type i32*. Constraint that is generated will be of the following 
form,

\begin{lstlisting}
    { lhs-id } = & { rhs-id }
\end{lstlisting}
\noindent
where \texttt{lhs-id} and \texttt{rhs-id} are the id's of the variables. 
\texttt{lhs-id} represents the \texttt{\%a} and \texttt{rhs-id} represents the 
node whose address is being assigned. In LLVM, this is a virtual node. So, 
library creates a dummy node.
Example,

\begin{lstlisting}
    1 = & 0
\end{lstlisting}


\subsubsection{load instruction}
Only those load instructions whose return type is a pointer type are used for
generating constraint. For e.g for the following load instruction

\begin{lstlisting}
    %a = load i32*, i32** %b
\end{lstlisting}

\noindent
RHS node is \%b and LHS node is \%a. And the type of instruction is load.

\subsubsection{store instruction}
In case of store instruction, only those instructions are considered in which
there is write to a pointer value. For e.g.

\begin{lstlisting}
    store i32* %a, i32** %b
\end{lstlisting}
\noindent
and the following store instruction will not be considered for constraint

\begin{lstlisting}
    store i32 10, i32* %a
\end{lstlisting}

In the first example, RHS node is \%a and LHS node is \%b. And the type of 
instruction is store.

\subsubsection{getelementptr instruction}
getelementptr instruction in LLVM is used for calculating the address 
of a member elements of a aggregate type. So it takes the base address of
the element of an aggregate type (e.g. struct or array) as the first argument.
Indices are provided in the form of comma separated list following the base 
address.
e.g.

\begin{lstlisting}
    %a = getelementptr { i32*, i32 }, { i32*, i32 }* %b, i32 0, i32 0
\end{lstlisting}

getelementptr instruction generates special kind of constraint in which LHS is 
same as \%a but RHS is a special node consisting of list of nodes being used as 
arguments. So, LHS here is \%a, and RHS is a node consisting of \%b, `i32 0' and 
`i32 0' in its use list. Or we can say that RHS uses \%b, i32 0 and i32 0. And 
the type of instruction is copy.

\subsubsection{phi instruction}
Phi instruction maintains the list of all the incoming values in its use list
as its RHS. While LHS is same as in LLVM statement.
E.g.
\begin{lstlisting}
    %a = phi i32* [ %0, %one ], [ %1, %two ], [ %3, %three ]
\end{lstlisting}

Phi instruction generates multiple constraints. Each of these constraint 
consists of single possible assignment to the LHS. So the constraints generated 
for above instruction will be, \%a = \%0, \%a = \%1, \%a = \%3. Each of these 
constraints is a copy assignment.

\subsubsection{bitcast instruction}
bitcast instruction can convert a pointer of one type to another type. It is
essentially a copy operation. LHS be the new pointer and RHS will be the old 
pointer.

E.g.
\begin{lstlisting}
    %a = bitcast i32* %b to i64*
\end{lstlisting}

\subsubsection{select instruction}
select instruction is similar to br instruction. So, instead of jumping, it
selects a value from a pair based on a condition variable. It also generates 
multiple constraints where each constraint is a possible assignment. Each 
constraint represents a copy assignment.

E.g.
\begin{lstlisting}
    %c = select i1 %cond, i32* %a, i32* %b
\end{lstlisting}

Constraints generated for this instruction will be, \%c = \%a \& \%c = \%b

\subsubsection{extractvalue instruction}
ExtractValue instruction is used to extract the value of a member element from 
an object of an aggregate type (struct or array). Whereas getelementptr is used 
to calculate the address of a member element with respect to the base address, 
the extractvalue instruction is used to directly access the value having a 
member element. It takes the object directly as the first argument and next 
arguments are the list of indices representing the index of the member 
elements of an aggregate type.

For e.g. if \%a is of type { i32*, i32 }, then 
following instruction will extract the address being holded by first member 
which is a pointer.

\begin{lstlisting}
 %first = extractvalue { i32*, i32 } %a, 0
\end{lstlisting}

Constraint generated is similar to the that of getelementptr instruction but 
instead it is of type copy. RHS contains the \%a node and in its use list, it 
contains the indices directly. So, unlike getelementptr, it doesn't store the 
id's of the node being used because indices are constants here. But in 
getelementptr instruction, indices can be variable too.

\subsubsection{insertvalue instruction}
Insertvalue instruction is used to insert a value into the member element of an 
object of an aggregate type.
The first operand of an ‘insertvalue‘ instruction is a value of struct or array 
type. The second operand is a first-class value to insert. The following 
operands are constant indices indicating the position at which to insert the 
value in a similar manner as indices in a ‘extractvalue‘ instruction. The value 
to insert must have the same type as the value identified by the indices.

For e.g.,

\begin{lstlisting}
 %aggregate = insertvalue { i32*, i32 } %a, i32* %b, 0
\end{lstlisting}

The above code is similar to the following C code,

\begin{lstlisting}
 t = a;
 t.first = b;
 aggregate = t;
\end{lstlisting}

The constraint generated is of type copy as it copies the address stored in \%b 
to a member of an object of aggregate type. RHS node of the constraint will be 
same as that of insertvalue instruction.

\section{Instructions not handled}
Following instructions are not being handled by the library,

\begin{itemize}
    \item call
    \item ret
    \item inttoptr
    \item ptrtoint
\end{itemize}

